ODGI: understanding pangenome graphs

### 2021-10-27 manuscript/main.tex
replace some more lost footnotes

Pangenome graphs stand to become a ubiquitous tool in genomics \citep{Eizenga_2020}. With ODGI we implemented a state-of-the-art tool suite that can transform, analyse, simplify, validate, and visualise pangenome graphs at large scale.
%
Lifting over annotations and linearizing nested graph structures place the suite as the bridge between traditional linear reference genome analysis and pangenome graphs. ODGI is a unique set of tools that enables scientists to explore and discover the underlying biology of pangenome graphs. Already, the tools are the backbone of pipelines such as the Pangenome Graph Builder\footnote{\url{https://github.com/pangenome/pggb} (accessed Oct 2021)} (PGGB) or nf-core/pangenome\footnote{\url{https://github.com/nf-core/pangenome} (accessed Oct 2021)}. Future work will add support for RNA and protein sequences and expand on metadata capabilities of large pangenome graphs.

@@

ODGI's visualization tools produce static, large-scale snapshots. Recent interactive browsers are reference centric \citep{Beyer2019, Yokoyama2019, Durant2021, Liang2021} or focusing on one dimension \citep{Wick_2015, Gonnella2018}. However, an interactive visualization solution that combines the 1D and 2D layout of a graph with annotation and read mapping information across different zoom levels is still missing. ODGI's graph layout algorithms could become a backbone of such tools, as is already the case for \textit{gfaestus}\footnote{\url{https://github.com/chfi/gfaestus}}, which currently relies on the \textit{odgi layout} output for its interactive 2D visualization.

>>>>

Pangenome graphs stand to become a ubiquitous tool in genomics \citep{Eizenga_2020}. With ODGI we implemented a state-of-the-art tool suite that can transform, analyse, simplify, validate, and visualise pangenome graphs at large scale.
%
Lifting over annotations and linearizing nested graph structures place the suite as the bridge between traditional linear reference genome analysis and pangenome graphs. ODGI is a unique set of tools that enables scientists to explore and discover the underlying biology of pangenome graphs. Already, the tools are the backbone of pipelines such as the Pangenome Graph Builder \citep{pggb} (PGGB) or nf-core/pangenome \citep{pangenome}. %\footnote{\url{https://github.com/nf-core/pangenome} (accessed Oct 2021)}. 
Future work will add support for RNA and protein sequences and expand on metadata capabilities of large pangenome graphs.

@@

ODGI's visualization tools produce static, large-scale snapshots. Recent interactive browsers are reference centric \citep{Beyer2019, Yokoyama2019, Durant2021, Liang2021} or focusing on one dimension \citep{Wick_2015, Gonnella2018}. However, an interactive visualization solution that combines the 1D and 2D layout of a graph with annotation and read mapping information across different zoom levels is still missing. ODGI's graph layout algorithms could become a backbone of such tools, as is already the case for \textit{gfaestus} \citep{gfaestus}%\footnote{\url{https://github.com/chfi/gfaestus}}, which currently relies on the \textit{odgi layout} output for its interactive 2D visualization.
